---
title: "hw04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

The goal of this homework is to solidify data wrangling skills by working some realistic problems in the grey area between data aggregation and data reshaping.

This file is organised by R functions, you could find:

- Data reshaping(Task 1)
  - [spread](#spread)
  - [gather](#gather)
- Joins(Task 2) 
  - [left_join](#left_join)
  - [right_join](#right_join)
  - [inner_join](#inner_join)
  - [full_join](#full_join)
  - [semi_join](#semi_join)
  - [anti_join](#anti_join)
  - [merge](#merge)
  - [match](#match)

Let's start with loading the libraries.
```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
```


## Data Reshaping(and relationship to aggregation)

### SPREAD
> Spread a key-value pair across multiple columns.

This is the activity 2 of the first task:

> Make a tibble with one row per year and columns for life expectancy for two or more countries.
>
>   - Use knitr::kable() to make this table look pretty in your rendered homework.
>
>   - Take advantage of this new data shape to scatterplot life expectancy for one country against that of another. 

In terms of the description, Canada and Australia are selected and compared here.

```{r}
df <- gapminder %>% 
  select(year, country, lifeExp) %>% 
  filter(country %in% c("Canada", "Australia")) %>%  # filter the data from both Canada and Australia
  spread(country, lifeExp) # spread the selected countries as columns

knitr::kable(df, caption = 'Life Expectancy by Year', digits = 2)
```

The new data shape is better for a scatterplot with the life expectancy of two countries since they are in the seperate columns now.

```{r}
df %>% 
  ggplot(aes(x = Australia, y = Canada, color = Canada>Australia)) + # colored by comparison
  geom_point(alpha = 0.8) +
  geom_text(aes(label=year)) + # show the corresponding years of the points
  theme_bw() +
  geom_abline(slope = 1) + # add a reference line
  ggtitle("Life expectancy of Canada against Australia over time")
```

To make the comparison clear, I used two diffenent colors and a reference line with the slope equals to 1. We could see that the life expectancy of two countries are very similar and Canada is a bit ahead in the history overall. 
 
### GATHER
> Gather takes multiple columns and collapses into key-value pairs, duplicating all other columns as needed. 

Use the `df` from the above section to show how `gather` works. 
```{r}
df <- gather(df, country, lifeExp, -c(year)) # gather Canada and Australia

knitr::kable(df)
```

With our new data shape, it's good to show how they spread respectively.
```{r}
df %>% 
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point() + 
  theme_bw() +
  ggtitle("Life expectancy of Canada against Australia over time")
  
```

## Join Prompts (join, merge, look up)

Create two data sources `my_gap` and `df` first. One is part of the `gapminder`, top 3 lifeExp of each continent. The other is consists of continent and hemisphere.
```{r}
df <- tibble(
  continent = c("Asia", "Oceania", "Europe", "Americas", "Antarctica"),
  hemisphere = c("northern", "southern", "northern", "northern","southern")
)
my_gap <- gapminder %>% 
  group_by(continent) %>% 
  top_n(3, lifeExp) %>%  
  select(continent, lifeExp)
```

Two tables are shown as follow:
```{r}
suppressPackageStartupMessages(library(gridExtra))
grid.arrange(tableGrob(my_gap), tableGrob(df), ncol = 2)
```

### LEFT_JOIN
>left_join(x, y): Return all rows from x, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. This is a mutating join.

```{r}
knitr::kable(left_join(my_gap, df))
```

Africa doesn't appear in `df`, so it has an `NA` for `hemisphere`.

### RIGHT_JOIN
>right_join(x, y): Return all rows from y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. This is a mutating join.

```{r}
knitr::kable(right_join(my_gap, df))
```

Antarctica doesn't appear in `my_gap`, so it has an `NA` for `lifeExp`.

### INNER_JOIN
> inner_join(x, y): Return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. This is a mutating join.

```{r}
knitr::kable(inner_join(my_gap, df))
```

According to the matching values from two dataframes, both Antarctica and Africa don't appear in the table.

### FULL_JOIN
> full_join(x, y): Return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing. This is a mutating join.

```{r}
knitr::kable(full_join(my_gap, df))
```

All continents appear in the table since full_join doesn't match values.

### SEMI_JOIN
> semi_join(x, y): Return all rows from x where there are matching values in y, keeping just columns from x. A semi join differs from an inner join because an inner join will return one row of x for each matching row of y, where a semi join will never duplicate rows of x. This is a filtering join.

```{r}
knitr::kable(semi_join(my_gap, df))
```

The rows of `semi_join` is the same with that of `inner_join`. But there is not "hemisphere" since it just keeps the columns from `my_gap`.

### ANTI_JOIN
> anti_join(x, y): Return all rows from x where there are not matching values in y, keeping just columns from x. This is a filtering join.

```{r}
knitr::kable(anti_join(my_gap, df))
```

It can be considered as`my_gap` deduct the output of `semi_join` above.

### MERGE
`merge` is a base R function and mainly based on SQL joins. In this section I would compare it with `dplyr` joins.

Both `merge` and `dplyr` joins could implemente different kinds of joins. The main difference between them is the intuitive syntax. As shown above, there are separate functions for each type of join in `dplyr`, whereas `merge` function differentiate them by arguments.

In the following examples, I would show how to make joins with `merge`.

- `left_join`
```{r}
knitr::kable(merge(my_gap, df, all.y = TRUE))
```

- `right_join`
```{r}
knitr::kable(merge(my_gap, df, all.y = TRUE))
```

- `full_join`
```{r}
knitr::kable(merge(my_gap, df, all = TRUE))
```

-`inner_join`
```{r}
knitr::kable(merge(my_gap, df))
```

All the outputs are the same with those of `dplyr` joins except the row order. It's worth mentioning that row order is guaranteed to be preserved in `dylyr` joins.

### MATCH

`Match` returns a vector of the positions of (first) matches of its first argument in its second. It does not perform any data join, but I would make a simple example to show how it works in an indirect way.

```{r}
pos <- match(df$continent, my_gap$continent) # the vector of the positions in my_gap
knitr::kable(my_gap[pos,])
```

From the matched positions, we could find the continents which appear in both `df` and `my_gap`.
